Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-24 paper.tex
...

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}

@@

    // 2. plays a "Bomb!" sound.
    await game.dt until s.sprite.frame == 10;
    {play_sound("plop")};

    // 3. particles, terrain, explosion sprite
    await game.dt until s.sprite.frame == 13;
    spawn PinguParticles(<...>) in particles;
    call Game_Remove({&bomber_radius}, <...>);
    do
      <...>
      spawn Sprite(<...>);      // explosion

      // 4. tick: hides the explosion sprite
      await game.dt;
    end
    await FOREVER;
  end

@@

Figure~\ref{lst.bomber} compares the implementations in C++ and \CEU.

In C++, the class \code{Bomber} defines the callbacks \code{draw} and
\code{update} to manage the state machine described above.
%
The class first defines one state variable for each action to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) first updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
switches to \emph{state-2} (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
43--44).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 34--35).
%
Note that a single numeric state variable suffices to track the states, but the
original authors probably chose to encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., \emph{methods + mutable state}).
%
Like double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality.

The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
variables and reflects the sequential state machine implicitly, using
\code{await} statements in direct style to separate the actions.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code/await} abstractions and execute in separate through the \code{spawn}
primitive (ln. 4--5).
The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every frame.
%
The code tracks the animation aliveness (ln. 7--27) and, on termination,
performs the last bomber action (ln. 30).
As soon as the animation starts, the code performs the first action (ln. 9).
The intermediate actions are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all actions happen in a contiguous
chunk of code (ln. 5--30) which handles no extra functionality.

@@

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%
Figure~\ref{lst.story} compares the implementations in C++ and \CEU.

In C++, the class \code{StoryScreenComponent} implements the method
\code{next\_text}, which is a callback for clicks in \code{>>>}.
%
The variable `pages` (ln. 4--5, 24--26) is a vector holding each page, but
which also encodes \emph{continuations} for the story progress:
each call to \code{next\_text} that advances the story (ln. 23--32) removes the 
current page (ln. 24) and sets the next action to perform (i.e., ``display a
new page'') in the variable \code{current\_page} (ln. 26).
Figure~\ref{fig.story} illustrates the continuation mechanism to advance 
pages and also a state machine for fast forwarding words (inside the dashed
rectangle).
The state variable \code{displayed} (ln. 6,15,20,21,27) switches between the
behaviors ``advancing text'' and ``advancing pages'', which are both handled
intermixed inside the method \code{next\_text}.

The code in \CEU uses the internal event \code{next\_text}, which is emitted
from clicks in \code{>>>}.
%
The sequential navigation from page to page uses a loop in direct style
(ln. 6--15) instead of explicit state variables for the continuation and state
machine.
While the text advances in an inner loop (hidden in ln. 9), we watch the
\code{next\_text} event that fast forwards it.
The loop may also eventually terminate with the time elapsing normally.
This way, we do not need a variable (such as `displayed` in C++) to switch 
between the states ``advancing text'' and ``advancing pages''.
The \code{par/or} makes the page advance logic to execute in parallel with the
redrawing code (ln. 13).
Whenever the page advances, the redrawing code is automatically aborted
(due to the \code{or} modifier).
The \code{await next\_text} in sequence (ln. 11) is the condition to advance to
the next page.
%
Note that, unlike the implementation in C++, the ``advancing text'' behavior is
not intermixed with the ``advancing pages'' behavior, instead, it is
encapsulated inside the inner loop nested with a deeper indentation (ln. 9).

@@

\subsubsection{Case Study: \emph{Bomber Action} \code{draw} and \code{update} Dispatching}

TODO

Figure~\ref{lst.hier} compares the implementations in C++ and \CEU.

@@

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.bomber.fsm}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

In \CEU, the \code{Bomber} action spawns a \code{Sprite} animation instance on
its body.
%
The \code{Sprite} instance (ln. 3) can react directly to external \code{dt} and
\code{redraw} events (which are analogous to \code{update} and \code{redraw}
callbacks, respectively), bypassing the program hierarchy entirely.
While and \emph{only while} the bomber abstraction is alive, the sprite
animation is also alive.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.54\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
GameSession::GameSession(<...>) :
    <...>
{
    <...>      // these widgets are always active...
    btpanel  = new ButtonPanel(<...>);
    pcounter = new PingusCounter(<...>);
    smallmap = new SmallMap(<...>);
    <...>
    uimgr->add(btpanel);  // ...but are added
    uimgr->add(pcounter); //  dynamically to the
    uimgr->add(smallmap); //  dispatching hierarchy
    <...>
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.46\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...> // other coexisting functionality
    spawn ButtonPanel(<...>);
    spawn PingusCounter(<...>);
    spawn SmallMap(<...>);
    <...> // other coexisting functionality
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the UI widgets lifecycle.
\label{lst.pingus}
}
\end{figure*}

Figure~\ref{fig.ui} shows the game UI widgets holding the buttons, pingus
counter, and small map which coexist with the game screen during its whole
lifespan.
%
Figure~\ref{lst.ui} compares the implementations in C++ and \CEU.

In C++, the widgets are created in the constructor of the class
\code{GameSession} (ln. 5--7), added to a UI container (ln. 9--11), and are
never removed since they must always be visible.
Arguably, to better express the intent of making them coexist with the game
screen, the widgets could be declared as top-level automatic (non-dynamic)
members.
However, the class uses a container to automate \code{draw} and \code{update}
dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
In turn, the container method \code{add} expects dynamically allocated children
only because they are automatically deallocated inside the container
destructor.
%
The dynamic nature of containers in C++ demand extra caution from programmers:
%
\begin{itemize}
\item When containers are part of a dispatching chain, it gets even harder to
      know which objects are dispatched:
      one has to ``simulate'' the program execution and track calls to
      \code{add} and \code{remove}.
\item For objects with dynamic lifespan, calls to \code{add} must always have
      matching calls to \code{remove}:
      missing calls to \code{remove} lead to memory and CPU leaks (to be
      discussed as the \emph{lapsed listener} problem in
      Section~\ref{sec.pats.lifespan.2}).
\end{itemize}

In Céu, entities that coexist just have to be created in the same lexical
block.
%
Since abstractions can react independently, they do not require a dispatching
container.
%
Lexical lifespan never requires containers, allocation and deallocation, or
explicit references.
In addition, all required memory is known at compile time, similarly to
stack-allocated local variables.
%
The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
lexical scope to delimit the lifespan of the explosion sprite to a single
frame.

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{github.com/an000/p/blob/master/README.md#5} }.
%
Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.

@@

code/await Pingu (<...>) do
    <...>
    loop do
        await game.dt;
        if Pingu_Is_Out_Of_Screen() then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the pingus lifecycle.
\label{lst.pingus}
}
\end{figure*}

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
also occurs in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

>>>>

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define,public,private},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\Code}[1] {{\texttt{#1}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}

@@

    // 2. plays a "Bomb!" sound.
    await game.update until s.sprite.frame == 10;
    {play_sound("plop")};

    // 3. particles, terrain, explosion sprite
    await game.update until s.sprite.frame == 13;
    spawn PinguParticles(<...>) in particles;
    call Game_Remove({&bomber_radius}, <...>);
    do
      <...>
      spawn Sprite(<...>);      // explosion

      // 4. tick: hides the explosion sprite
      await game.update;
    end
    await FOREVER;
  end

@@

In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
callbacks \code{draw} and \code{update} to manage the state machine described
above.
%
The class first defines one state variable for each action to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) first updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
switches to \emph{state-2} (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
43--44).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 34--35).
%
Note that a single numeric state variable suffices to track the states, but the
original authors probably chose to encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., \emph{methods + mutable state}).
%
Like double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality.

The equivalent code in \CEU for the \emph{Bomber action} in
Figure~\ref{lst.bomber}.b doesn't require state variables and reflects the
sequential state machine implicitly, using \code{await} statements in direct
style to separate the actions.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code/await} abstractions and execute in separate through the \code{spawn}
primitive (ln. 4--5).
The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every frame.
%
The code tracks the animation aliveness (ln. 7--27) and, on termination,
performs the last bomber action (ln. 30).
As soon as the animation starts, the code performs the first action (ln. 9).
The intermediate actions are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all actions happen in a contiguous
chunk of code (ln. 5--30) which handles no extra functionality.

@@

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%

In C++, the class \code{StoryScreenComponent} in Figure~\ref{lst.story}.a
implements the method \code{next\_text}, which is a callback for clicks in
\code{>>>}.
%
The variable `pages` (ln. 4--5, 24--26) is a vector holding each page, but
which also encodes \emph{continuations} for the story progress:
each call to \code{next\_text} that advances the story (ln. 23--32) removes the 
current page (ln. 24) and sets the next action to perform (i.e., ``display a
new page'') in the variable \code{current\_page} (ln. 26).
Figure~\ref{fig.story} illustrates the continuation mechanism to advance 
pages and also a state machine for fast forwarding words (inside the dashed
rectangle).
The state variable \code{displayed} (ln. 6,15,20,21,27) switches between the
behaviors ``advancing text'' and ``advancing pages'', which are both handled
intermixed inside the method \code{next\_text}.

The code in \CEU in Figure~\ref{lst.story}.b uses the internal event
\code{next\_text}, which is emitted from clicks in \code{>>>}.
%
The sequential navigation from page to page uses a loop in direct style
(ln. 6--15) instead of explicit state variables for the continuation and state
machine.
While the text advances in an inner loop (hidden in ln. 9), we watch the
\code{next\_text} event that fast forwards it.
The loop may also eventually terminate with the time elapsing normally.
This way, we do not need a variable (such as `displayed` in C++) to switch 
between the states ``advancing text'' and ``advancing pages''.
The \code{par/or} makes the page advance logic to execute in parallel with the
redrawing code (ln. 13).
Whenever the page advances, the redrawing code is automatically aborted
(due to the \code{or} modifier).
The \code{await next\_text} in sequence (ln. 11) is the condition to advance to
the next page.
%
Note that, unlike the implementation in C++, the ``advancing text'' behavior is
not intermixed with the ``advancing pages'' behavior, instead, it is
encapsulated inside the inner loop nested with a deeper indentation (ln. 9).

@@

\subsubsection{Case Study: \emph{Bomber Action} \Code{draw} and \Code{update} Dispatching}

@@

In C++, the class \code{Bomber} presented in Figure~\ref{lst.hier}.a declares a
\code{sprite} member to handle its animation frames
(Figure~\ref{fig.bomber.fsm}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
spawns a \code{Sprite} animation instance on its body.
%
The \code{Sprite} instance (ln. 3) can react directly to external \code{update}
and \code{draw} events, bypassing the program hierarchy entirely.
While and \emph{only while} the bomber abstraction is alive, the sprite
animation is also alive.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.54\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
GameSession::GameSession(<...>) :
    <...>
{
    <...>      // these widgets are always active...
    btpanel  = new ButtonPanel(<...>);
    pcounter = new PingusCounter(<...>);
    smallmap = new SmallMap(<...>);
    <...>
    uimgr->add(btpanel);  // ...but are added
    uimgr->add(pcounter); //  dynamically to the
    uimgr->add(smallmap); //  dispatching hierarchy
    <...>
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.46\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...> // other coexisting functionality
    spawn ButtonPanel(<...>);
    spawn PingusCounter(<...>);
    spawn SmallMap(<...>);
    <...> // other coexisting functionality
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the UI widgets lifecycle.
\label{lst.ui}
}
\end{figure*}

Figure~\ref{fig.ui} shows the game UI widgets with action buttons, the pingus
counter, and a small map which all coexist with the game screen during its
whole lifespan.
%
In C++, the widgets are created in the constructor of the class
\code{GameSession} in Figure~\ref{lst.ui}.a (ln. 5--7), added to a UI container
(ln. 9--11), and are never removed since they must always be visible.
Arguably, to better express the intent of making them coexist with the game
screen, the widgets could alternatively be declared as top-level automatic
(non-dynamic) members.
However, the class uses a container to automate \code{draw} and \code{update}
dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
In turn, the container method \code{add} only expects dynamically allocated
children because they are automatically deallocated inside the container
destructor.
%
The dynamic nature of containers in C++ demand extra caution from programmers:
%
\begin{itemize}
\item When containers are part of a dispatching chain, it gets even harder to
      know which objects are dispatched at a given moment:
      one has to ``simulate'' the program execution and track calls to
      \code{add} and \code{remove}.
\item For objects with dynamic lifespan, calls to \code{add} must always have
      matching calls to \code{remove}:
      missing calls to \code{remove} lead to memory and CPU leaks (to be
      discussed as the \emph{lapsed listener} problem in
      Section~\ref{sec.pats.lifespan.2}).
\end{itemize}

In Céu, entities that coexist just have to be created in the same lexical
block in Figure~\ref{lst.ui}.b (ln. 3--5).
%
Since abstractions can react independently, they do not require a dispatching
container.
%
Lexical lifespan never requires containers, allocation and deallocation, or
explicit references.
In addition, all required memory is known at compile time, similarly to
stack-allocated local variables.
%
The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
lexical scope to delimit the lifespan of the explosion sprite to a single
frame.

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{github.com/an000/p/blob/master/README.md#5} }.
%

@@

code/await Pingu (<...>) do
    <...>
    loop do
        await game.update;
        if Pingu_Is_Out_Of_Screen() then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the pingus lifecycle.
\label{lst.pingus}
}
\end{figure*}

In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
container that holds all pingus alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{draw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
also occurs in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen in Figure~\ref{lst.pingus}.b spawns a new \code{Pingu} on every
invocation of \code{Pingu\_Spawn} (ln. 4--7).
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.
